---
title: "ST558 - Project 1"
author: "Kayla Nowak"
format: html
---

## Introduction

This project builds a set of generalized queries for accessing and summarizing data from the Public Use Microdata Sample (PUMS). PUMS is part of the US Census Bureau's annual American Community Survey.

-   Data Source: <https://www.census.gov/data/developers/data-sets/census-microdata-api.html>
-   Variables (from 2024): <https://api.census.gov/data/2024/acs/acs1/pums/variables.html>

This project was designed to access the 1-Year PUMS data sets for years 2021 through 2024 only. Additionally, PUMS data for Puerto Rico are not included the scope of this project.

## Data Processing Functions

Load essential packages and set API key.

```{r}
#| message: False
#| warning: False

library(tidyverse)
library(httr)
library(jsonlite)
library(ggplot2)
```

Core R objects, used throughout.

```{r}
my_key <- "07a32887514d391e8f71d7c20274286c27064b18"

all_years <- c("2021", "2022", "2023", "2024")

all_nums <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
all_cats <- c("FER", "HHL", "SCH", "SCHL", "SEX")
all_vars <- c("PWGTP", all_nums, all_cats, "ST", "STATE", "REGION", "DIVISION")
```

### Variable Values

Pull and parse variable names and values for the PUMS data in a given year.

-   Input: `year` (2021, 2022, 2023, 2024)
-   Return: `data/vars_yyyy.rda` (as list)

Function:

```{r}

getPUMSVars <- function(year){
  
  yearc <- as.character(year)

  ## Define variables of interest
  if (yearc %in% c("2021", "2022")) {
    want <- all_vars[-14]
  } else if (yearc %in% c("2023", "2024")) {
    want <- all_vars[-13]
  } else {
    stop("Invalid year input. Function only accepts 2021-2024")
  }
  
  ## Pull + Parse JSON
  url <- paste0("https://api.census.gov/data/",yearc,"/acs/acs1/pums/variables.json")
  pulled  <- GET(url)
  parsed  <- fromJSON(rawToChar(pulled$content))
  
  ## Subset List to Variables of Interest
  vars <- parsed$variables[want]
  
  ## Save to /data as .rds
  saveRDS(vars, file= paste0("data/vars",yearc,".rds"))
}

```

Called for all valid years, then commented out.

*Note: This code only needs to be run once since the results are permanently stored under in the repository under /data. Calls have been commented out for computational efficiency of future QMD executions.*

```{r}
#getPUMSVars(2021)
#getPUMSVars(2022)
#getPUMSVars(2023)
#getPUMSVars(2024)
```

### API Pull: Helpers

The following helper functions perform dedicated tasks and are called within the main, wrapper function (`getPUMSData` defined later).

`getFromAPI()` queries the Census API for PUMAS data and returns a matrix of the requested data.

-   Input: query parameters, as given to the main/wrapper function.
-   Return: a matrix

```{r}
getFromAPI <- function(year, num, cat, geo, subgeo){
  
  q <- paste(cat, "PWGTP", num, sep=",") 
  url <- paste0("https://api.census.gov/data/",as.character(year),
                "/acs/acs1/pums?get=",q,
                "&for=",geo,
                ":*&SCHL=24&key=", my_key)
  
  pulled <- GET(url)
  parsed <- fromJSON(rawToChar(pulled$content))
  
  ##REMOVE SCHL=24 later
  
}
```

`makeTidy()` converts the requested data to a tibble with meaningful column names.

-   Input: a matrix generated by `getFromAPI`
-   Return: a tibble

```{r}
makeTidy <- function(parsed){
  
  #Get colnames from 1st row
  colnames(parsed) <- parsed[1,]
  
  #Make parsed tidy with colnames
  tdata  <- as_tibble(parsed) |> slice(-1) |>
    mutate(PWGTP=as.numeric(PWGTP))
}
```

`cleanCat()` converts the requested categorical variable to a factor, based on the relevant data dictionary stored in /data.

-   Input: a tibble generated by makeTidy, the data dictionary, and the cat user input
-   Return: an updated tibble

```{r}
cleanCat <- function(tdata,dictionary,cat){
  
  #Store variable labels
  lookup <- unlist(dictionary[[cat]]$values$item)
  
  #Apply labels and levels
  if(cat=="SEX") {
    tdata <- tdata |> mutate(SEX = factor(SEX, levels=names(lookup), labels=lookup))
  } else if(cat=="FER"){
    tdata <- tdata |> mutate(FER = factor(FER, levels=names(lookup), labels=lookup))
  } else if(cat=="HHL"){
    tdata <- tdata |> mutate(HHL = factor(HHL, levels=names(lookup), labels=lookup))
  } else if(cat=="SCH"){
    tdata <- tdata |> mutate(SCH = factor(SCH, levels=names(lookup), labels=lookup))
  } else if(cat=="SCHL"){
    tdata <- tdata |> mutate(SCHL = factor(SCHL, levels=names(lookup), labels=lookup))
  } 

}
```

`cleanNum()` converts the requested 2nd numeric variable to a number or derived midpoint time, based on the relevant data dictionary stored in /data.

-   Input: a tibble generated by makeTidy, the data dictionary, and the num user input
-   Return: an updated tibble

```{r}
cleanNum <- function(tdata,dictionary,num){
  
  #Store variable labels
  lookup <- unlist(dictionary[[num]]$values$item)
  
  #Apply labels and levels
  if(num=="AGEP") {
    tdata <- tdata |> mutate(AGEP = as.numeric(AGEP))
  } else if(num=="GASP"){
    tdata <- tdata |> mutate(GASP = as.numeric(GASP), 
                             GASP = ifelse(GASP==3,NA,GASP)) #set coded missing to zero
  } else if(num=="GRPIP"){
    tdata <- tdata |> mutate(GRPIP = as.numeric(GRPIP))
  } else if(num=="JWAP"){
    tdata <- tdata |> mutate(JWAP = factor(JWAP, levels=names(lookup), labels=lookup))
  } else if(num=="JWDP"){
    tdata <- tdata |> mutate(JWDP = factor(JWDP, levels=names(lookup), labels=lookup))
  } else if(num=="JWMNP") {
    tdata <- tdata |> mutate(JWMNP = as.numeric(JWMNP))
  } 
 
  ##COME BACK FOR TIME CLEANING! 
}
```

`cleanGeo()` converts the requested geographic variable to a factor based on the relevant data dictionary stored in /data.

-   Input: a tibble generated by makeTidy, the data dictionary, and the geo user input
-   Return: an updated tibble

```{r}
cleanGeo <- function(tdata,dictionary,geo){
  
  #Store variable labels
  lookup <- unlist(dictionary[[geo]]$values$item)
  
  #Apply labels and levels
  if(geo=="STATE") {
    tdata <- tdata |> mutate(state = factor(state, levels=names(lookup), labels=lookup))
  } else if(geo=="ST"){
    tdata <- tdata |> mutate(st = factor(st, levels=names(lookup), labels=lookup))
  } else if(geo=="REGION"){
    tdata <- tdata |> mutate(region = factor(region, levels=names(lookup), labels=lookup))
  } else if(geo=="DIVISION"){
    tdata <- tdata |> mutate(division = factor(division, levels=names(lookup), labels=lookup))
  }

}
```

### API Pull: Main

The `getPUMSData` function ...

-   `year` accepts 2021, 2022, 2023, 2024
-   `nums` accepts one numeric variable: AGEP (default), GASP, GRPIP, JWAP, JWDP, JWMNP
    -   Note: "PWGTP" is automatically included in addition to the requested variable
-   `cats` accepts one categorical variable: FER, HHL, SCH, SCHL, SEX (default).
-   `geo` accepts one level of geography: REGION, DIVISION, STATE (default)
    -   Note: either "STATE" or "ST" may be used to request state-level geography
-   `subgeo` accepts one sub-level of geography:


```{r}
getPUMSData <- function(year, num, cat="SEX", geo="STATE", subgeo="ALL", ...){
  
  ## CHECK FOR VALID INPUTS
  #--ADD THESE BACK LATER--
  
  ## RUN API PULL FUNCTION
  parsed <- getFromAPI(year, num, cat, geo, subgeo)

  ## RUN TIDY FUNCTION
  tdata  <- makeTidy(parsed)
  
  ## RUN CLEANING FUNCTIONS
  dictionary <- readRDS(paste0("data/vars",as.character(year),".rds"))
  
  tdata <- cleanCat(tdata,dictionary,cat)
  tdata <- cleanNum(tdata,dictionary,num)
  tdata <- cleanGeo(tdata,dictionary,geo)
  
}

```

### Testing

```{r}
wip <- getPUMSData(year=2024, num="GASP", cat="HHL", geo="STATE", subgeo="ALL")

#Check results
wip
str(wip)
wip |> count(GASP)
```



## --- HERE ---


```{r}
#| eval: false

input_checks <- function(year, 
                        nums=c("AGEP","PWGTP"), 
                        cats=c("SEX"), 
                        geo=c("STATE"), 
                        subgeo=NULL, ...) {
  
  ## YEAR
  yearc <- as.character(year)
  
  #Check validity
  if (!yearc %in% c("2021", "2022", "2023", "2024")) {
    stop("Invalid year input. Function only accepts 2021-2024.")
  }
  
  ## NUMS
  nums <- toupper(nums)
  
  #Check validity
  if (length(nums) > 2 | (length(nums)==2 & !any(nums == "PWGTP")) ){
    stop("Invalid NUMS input. Only 1 numeric variable in addition to PWGTP may be requested.")
    
  } else if (any(!nums %in% c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"))) {
    stop("Invalid NUMS input. The requested variable(s) are not on the approved numeric list.")
  } else if (length(nums==1) & nums == "PWGTP"){
    stop("Invalid NUMS input. At least 1 numeric variable other than PWGTP must be provided.")
  } else if (length(nums==1) & nums != "PWGTP"){
    #Automatically include PWGTP on a single-input request
    nums <- c(nums, "PWGTP")
  }
  
  ## CATS 
  cats <- toupper(cats)
  
  #Check validity
  if (length(cats) > 1){
    stop("Invalid CATS input. Only 1 categorical variable may be provided.")
    
  } else if (!cats %in% c("FER", "HHL", "SCH", "SCHL", "SEX")){
    stop("Invalid CATS input. The requested variable is not on the approved categorical list.")
    
  }

  ## GEO
  geo <- toupper(geo)
  
  #Check validity
  if (length(geo) > 1){
    stop("Invalid GEO input. Only 1 geography variable may be provided.")
    
  } else if (!geo %in% c("REGION", "DIVISION", "STATE", "ST")){
    stop("Invalid GEO input. The requested variable is not on the approved geography list.")
    
  } else {
    
    #Force "ST" vs. "STATE" if entered wrong.
    if(yearc %in% c("2021", "2022") & geo=="STATE") {
      geo <- "ST"
    } else if(yearc %in% c("2023", "2024") & geo=="ST") {
      geo <- "STATE"
    }
  }
  
  ## SUBGEO
  
  #Grab year-specific valid inputs
  vars <- readRDS(paste0("data/vars",yearc,".rds"))
  subR <- unlist(vars$REGION$values$item)
  subD <- unlist(vars$DIVISION$values$item)
  if(yearc %in% c("2021", "2022")) { 
    subS <- unlist(vars$ST$values$item)
  } else {
    subS <- unlist(vars$STATE$values$item) 
  }
  
  #Check validity
  if (length(subgeo) > 1){
    stop("Invalid SUBGEO input. Only 1 sub-geography variable may be provided.")
    
  } else if (geo=="REGION" & !subgeo %in% subR) {
    stop("Invalid SUBGEO input. The provided phrase is invalid for geo = REGION. Note: this field is case-sensitive.")
    
  } else if(geo=="DIVISION" & !subgeo %in% subD){
    stop("Invalid SUBGEO input. The provided phrase is invalid for geo = DIVISION Note: this field is case-sensitive.")
      
  } else if(geo %in% c("STATE", "ST") & !subgeo %in% subS){
    stop("Invalid SUBGEO input. The provided phrase is invalid for geo = STATE Note: this field is case-sensitive.")
      
  } else if (is.null(subgeo) & geo=="REGION") {
    subgeo <- "Northeast"  #set default 
    
  } else if(is.null(subgeo) & geo=="DIVISION") {
     subgeo <- "New England (Northeast region)"  #set default 
   
  } else if(is.null(subgeo) & geo %in% c("STATE", "ST")) {
     subgeo <- "District of Columbia/DC"  #set default 
   
  }
  
}

```

## Data Summarizing Functions

## Testing

### getPUMSVars()

Expected to fail:

```{r}
#| eval: false
#| error: true

getPUMSVars(2020)  #Invalid year
getPUMSVars()      #No Year
```

### getPUMSData()

Expected to work:

Expected to fail:

```{r}
#| eval: false
#| error: true

#Invalid Year
getPUMSData(2020, nums, cats, geo, subgeo)

#Invalid Nums
getPUMSData(2021, nums=c("AGEP", "GASP"))  #too many inputs
getPUMSData(2021, nums=c("PWGTP"))         #too few inputs
getPUMSData(2021, nums=c("STATE"))         #unapproved var
getPUMSData(2021, nums=c("STATE","PWGTP")) #unapproved var

#Invalid Cats
getPUMSData(2021, nums=c("AGEP"), cats=c("SEX","FER"))  #too many inputs
getPUMSData(2021, nums=c("AGEP"), cats=c("STATE"))      #unapproved var

#Invalid Geo
getPUMSData(2021, nums=c("AGEP"), geo=c("STATE","REGION"))  #too many inputs
getPUMSData(2021, nums=c("AGEP"), geo=c("STA"))             #unapproved var

#Invalid Subgeo
getPUMSData(2021, nums=c("AGEP"), geo=c("STATE"), subgeo=c("NC", "DC"))  #too many inputs
getPUMSData(2021, nums=c("AGEP"), geo=c("STATE"), subgeo=c("NC"))        #invalid inputs

```

## Investigation

**Topic**

**Results**

**Discussion**
