---
title: "ST558 - Project 1"
author: "Kayla Nowak"
format: html
---

## Introduction

This project builds a set of generalized queries for accessing and summarizing data from the Public Use Microdata Sample (PUMS). PUMS is part of the US Census Bureau's annual American Community Survey.

-   Data Source: <https://www.census.gov/data/developers/data-sets/census-microdata-api.html>
-   Variables (from 2024): <https://api.census.gov/data/2024/acs/acs1/pums/variables.html>

This project was designed to access the 1-Year PUMS data sets for years 2021 through 2024 only. Additionally, PUMS data for Puerto Rico are not included the scope of this project.

## Functions

Load essential packages and set API key.

```{r}
#| message: False
#| warning: False

library(tidyverse)
library(httr)
library(jsonlite)
library(ggplot2)

```

Core R objects, used throughout.

```{r}
my_key <- "07a32887514d391e8f71d7c20274286c27064b18"

all_years <- c("2021", "2022", "2023", "2024")

all_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER",
              "HHL", "SCH", "SCHL", "SEX", "ST", "STATE", "REGION", "DIVISION")

```



### 1. Data Processing

**Variable Values**

Pull and parse variable names and values for the PUMS data in a given year.

-   Input: `year` (2021, 2022, 2023, 2024)
-   Return: `data/vars_yyyy.rda` (as list)

Function:

```{r}

getPUMSVars <- function(year){
  
  yearc <- as.character(year)

  ## Define variables of interest
  if (yearc %in% c("2021", "2022")) {
    want <- all_vars[-14]
  } else if (yearc %in% c("2023", "2024")) {
    want <- all_vars[-13]
  } else {
    stop("Invalid year input. Function only accepts 2021-2024")
  }
  
  ## Pull + Parse JSON
  url <- paste0("https://api.census.gov/data/",yearc,"/acs/acs1/pums/variables.json")
  pulled  <- GET(url)
  parsed  <- fromJSON(rawToChar(pulled$content))
  
  ## Subset List to Variables of Interest
  vars <- parsed$variables[want]
  
  ## Save to /data as .rds
  saveRDS(vars, file= paste0("data/vars",yearc,".rds"))
}

```

Called for all valid years:

```{r}
#getPUMSVars(2021)
#getPUMSVars(2022)
#getPUMSVars(2023)
#getPUMSVars(2024)
```

**API Data Pull**


```{r}
yearc <- "2024"
nums  <- c("PWGTP", "AGEP", "STATE")

getPUMSData <- function(year, nums=c("AGEP","PWGTP"), cats, geo, region) {
  
  ## Check YEAR input
  yearc <- as.character(year)
  if (!yearc %in% c("2021", "2022", "2023", "2024")) {
    stop("Invalid year input. Function only accepts 2021-2024.")
  }
  
  ## Check NUMS input
  nums <- toupper(nums)
  if (length(nums) == 1 & nums == ("PWGTP")){
    #At least 1 numeric variable other than PWGTP required.
    stop("Invalid NUMS input. Must specify one more numeric variable in addition to PWGTP.")
    
  } else if (any(!nums %in% all_vars)){
    #Only variables on the approved list are accepted.
    #Note: any(not-in) == TRUE when any elements of nums are NOT in the elements of all_vars
    stop("Invalid NUMS input. At least one requested variable is not on the approved list.")
    
  } else if (any(!nums %in% c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"))) {
    #Only ~numeric~ variables on the approved list are accepted. Ask user to fix.
    stop("Invalid NUMS input. At least one requested variable is non-numeric.")
    
  } else {
    
    #Fix "ST" vs. "STATE" if entered wrong.
    if(yearc %in% c("2021", "2022") & nums %in% c("STATE")) {
      nums[which(nums == "STATE")] <- "ST"
      
    } else if(yearc %in% c("2023", "2024") & nums %in% c("ST")) {
      nums[which(nums == "ST")] <- "STATE"
    }
    
    #Ensure PWGTP is on list
    nums <- c(nums, "PWGTP")
    nums <- unique(nums)
  }
  
  ## Check CATS input
  
  ## Check GEO input
  
  ## Check REGION input
  
  ## Run API pull via helper function
  
}

```


### 2. Data Summarizing

## Testing

### With Results Returned

### With Errors/Stop Returned

## Investigation

**Topic**

**Results**

**Discussion**
