---
title: "ST558 - Project 1"
author: "Kayla Nowak"
format: html
---

# Introduction

This project builds a set of generalized queries for accessing and summarizing data from the Public Use Microdata Sample (PUMS). PUMS is part of the US Census Bureau's annual American Community Survey.

-   Data Source: <https://www.census.gov/data/developers/data-sets/census-microdata-api.html>
-   Variables (from 2024): <https://api.census.gov/data/2024/acs/acs1/pums/variables.html>

This project was designed to access the 1-Year PUMS data sets for years 2021 through 2024 only. Additionally, PUMS data for Puerto Rico are not included the scope of this project.

# Data Processing Functions

Load essential packages and set API key.

```{r}
#| message: False
#| warning: False

library(tidyverse)
library(jsonlite)
library(ggplot2)
```

Core R objects, used throughout.

```{r}
my_key <- "07a32887514d391e8f71d7c20274286c27064b18"

all_years <- c("2021", "2022", "2023", "2024")

all_nums <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
all_cats <- c("FER", "HHL", "SCH", "SCHL", "SEX")
all_vars <- c("PWGTP", all_nums, all_cats, "ST", "STATE", "REGION", "DIVISION")
```

## Variable Values

`getPUMSVars()` pulls and parses variable names and values for the PUMS data in a given year.

One year-specific .rda file is returned, and this file is saved in the repository's `/data` folder for future access. Later functions reference these .rda files when defining the `dictionary`.

-   Input: `year` (2021, 2022, 2023, 2024)
-   Return: `data/vars_yyyy.rda` (as list)

Function:

```{r}

getPUMSVars <- function(year){
  
  yearc <- as.character(year)

  ## Define variables of interest
  if (yearc %in% c("2021", "2022")) {
    want <- all_vars[-14]
  } else if (yearc %in% c("2023", "2024")) {
    want <- all_vars[-13]
  } else {
    stop("Invalid year input. Function only accepts 2021-2024")
  }
  
  ## Pull + Parse JSON
  url <- paste0("https://api.census.gov/data/",yearc,"/acs/acs1/pums/variables.json")
  pulled  <- httr::GET(url)
  parsed  <- fromJSON(rawToChar(pulled$content))
  
  ## Subset List to Variables of Interest
  vars <- parsed$variables[want]
  
  ## Save to /data as .rds
  saveRDS(vars, file= paste0("data/vars",yearc,".rds"))
}

```

Called for all valid years, then commented out.

*Note: This code only needs to be run once since the results are permanently stored under in the repository under /data. Calls have been commented out for computational efficiency of future QMD executions.*

```{r}
#getPUMSVars(2021)
#getPUMSVars(2022)
#getPUMSVars(2023)
#getPUMSVars(2024)
```

## API Pull: Helpers

The following helper functions perform dedicated tasks and are called within the main, wrapper function (`getPUMSData` defined later).

### `getFromAPI()`

This function queries the Census API for PUMAS data and returns a matrix of the requested data.

-   Input: all query parameters, as given to the main/wrapper function.
-   Return: a matrix

```{r}
getFromAPI <- function(year, num, cat, geo, subgeo,...){
  
  q <- paste(cat, "PWGTP", num, sep=",") 
  url <- paste0("https://api.census.gov/data/",as.character(year),
                "/acs/acs1/pums?get=",q,
                "&for=",geo,":",subgeo,
                "&SCHL=24&key=", my_key)
  
  pulled <- httr::GET(url)
  parsed <- fromJSON(rawToChar(pulled$content))
  
  ##REMOVE SCHL=24 later

}
```

### `makeTidy()`

This function converts the requested data to a tibble with meaningful column names.

-   Input: a matrix generated by `getFromAPI`
-   Return: a tibble

```{r}
makeTidy <- function(parsed,...){
  
  #Get colnames from 1st row
  colnames(parsed) <- parsed[1,]
  
  #Make parsed tidy with colnames
  tdata  <- as_tibble(parsed) |> slice(-1) |>
    mutate(PWGTP=as.numeric(PWGTP))
}
```

### `cleanCat()`

This function converts the requested categorical variable to a factor, based on the relevant data dictionary stored in `/data`.

-   Input: a tibble generated by `makeTidy`, the data dictionary, and the `cat` user input
-   Return: an updated tibble

```{r}
cleanCat <- function(tdata,dictionary,cat,...){
  
  #Store variable labels
  lookup <- unlist(dictionary[[cat]]$values$item)
  
  #Apply labels and levels
  if(cat=="SEX") {
    tdata <- tdata |> mutate(SEX = factor(SEX, levels=names(lookup), labels=lookup))
  } else if(cat=="FER"){
    tdata <- tdata |> mutate(FER = factor(FER, levels=names(lookup), labels=lookup))
  } else if(cat=="HHL"){
    tdata <- tdata |> mutate(HHL = factor(HHL, levels=names(lookup), labels=lookup))
  } else if(cat=="SCH"){
    tdata <- tdata |> mutate(SCH = factor(SCH, levels=names(lookup), labels=lookup))
  } else if(cat=="SCHL"){
    tdata <- tdata |> mutate(SCHL = factor(SCHL, levels=names(lookup), labels=lookup))
  } 

}
```

### `cleanTime()`

This is a helper function called within the `cleanNum` helper function (defined next). It converts the timing variables (JWAP and JWDP) from a time period (HH:MM to HH:MM) to a derived midpoint time. The resulting midpoint time is stored as R numeric data with the 12-hour time format (`I:M p`).

-   Input: the `num` user input; only use with JWAP or JWDP
-   Return: a vector

```{r}
cleanTime <- function(num,...){

  tvar <- as.character(num)
  
  #Split character times into vector with 2 elements
  split <- strsplit(tvar, " to ")[[1]]
  
  #Strip "." for lubridate
  split2 <- gsub("\\.", "", split)

  #Convert to numeric time with lubridate:
  nt1 <- lubridate::parse_date_time(split2[1], orders = "I:M p")
  nt2 <- lubridate::parse_date_time(split2[1], orders = "I:M p")
    
  #Calculate midpoint (numeric)
  mpt <- format(nt1 + (nt2-nt1)/2, "%I:%M %p")
  return(mpt)
  
}
```

### `cleanNum()`

This function converts the requested numeric variable to a number or derived midpoint time, based on the relevant data dictionary stored in `/data`.

-   Input: a tibble generated by `makeTidy`, the data dictionary, and the `num` user input
-   Return: an updated tibble

```{r}
cleanNum <- function(tdata,dictionary,num,...){
  
  #Store variable labels
  lookup <- unlist(dictionary[[num]]$values$item)
  
  #Apply labels and levels
  if(num=="AGEP") {
    tdata <- tdata |> mutate(AGEP = as.numeric(AGEP))
  } else if(num=="GASP"){
    tdata <- tdata |> mutate(GASP = as.numeric(GASP), 
                             GASP = ifelse(GASP==3,NA,GASP)) #set coded missing to zero
  } else if(num=="GRPIP"){
    tdata <- tdata |> mutate(GRPIP = as.numeric(GRPIP))
  } else if(num=="JWAP"){
    tdata <- tdata |> mutate(JWAP = factor(JWAP, levels=names(lookup), labels=lookup),
                             )#JWAP = cleanTime(JWAP))
  } else if(num=="JWDP"){
    tdata <- tdata |> mutate(JWDP = factor(JWDP, levels=names(lookup), labels=lookup),
                             )#JWDP = cleanTime(JWDP))
  } else if(num=="JWMNP") {
    tdata <- tdata |> mutate(JWMNP = as.numeric(JWMNP))
  } 

}
```

### `cleanGeo()`

This function converts the requested geographic variable to a factor based on the relevant data dictionary stored in `/data`.

-   Input: a tibble generated by `makeTidy`, the data dictionary, and the `geo` user input
-   Return: an updated tibble

```{r}
cleanGeo <- function(tdata,dictionary,geo,...){
  
  #Store variable labels
  lookup <- unlist(dictionary[[geo]]$values$item)
  
  #Apply labels and levels
  if(geo=="STATE") {
    tdata <- tdata |> mutate(state = factor(state, levels=names(lookup), labels=lookup))
  } else if(geo=="ST"){
    tdata <- tdata |> mutate(st = factor(st, levels=names(lookup), labels=lookup))
  } else if(geo=="REGION"){
    tdata <- tdata |> mutate(region = factor(region, levels=names(lookup), labels=lookup))
  } else if(geo=="DIVISION"){
    tdata <- tdata |> mutate(division = factor(division, levels=names(lookup), labels=lookup))
  }

}
```

## API Pull: Main

The **`getPUMSData`** function ...

-   `year` accepts 2021, 2022, 2023, 2024
-   `nums` accepts one numeric variable: AGEP (default), GASP, GRPIP, JWAP, JWDP, JWMNP
    -   Note: "PWGTP" is automatically included in addition to the requested variable
-   `cats` accepts one categorical variable: FER, HHL, SCH, SCHL, SEX (default).
-   `geo` accepts one level of geography: REGION, DIVISION, STATE (default)
    -   Note: either "STATE" or "ST" may be used to request state-level geography
-   `subgeo` accepts one sub-level of geography: NULL (default), ALL, or a numeric code
    -   The numeric code lists are specific to each level of geography. Please see the [online PUMAS variable documentation](https://api.census.gov/data/2024/acs/acs1/pums/variables.html) for a complete list.
    -   If no sub-level of geography is provided, the following defaults are used:
        -   REGION = 1 ("Northeast")
        -   DIVISION = 1 ("New England (Northeast region)")
        -   STATE = 11 ("District of Columbia/DC")

```{r}
getPUMSData <- function(year=2021, num="AGEP", cat="SEX", geo="STATE", subgeo=NULL,...){
 
  ## CHECK FOR VALID INPUTS
  
  #year:
  if (!as.character(year) %in% c("2021", "2022", "2023", "2024")){
    stop("Invalid YEAR input. Function only accepts 2021-2024.")
  } 
  
  #num:
  num <- toupper(num)
  
  if (length(num)>1) {
    stop("Invalid NUM input. Only 1 variable permitted - PWGTP is automatically included.")
  } else if(num=="PWGTP"){
    stop("Invalid NUM input. PWGTP is automatically included, a different numeric variable is required.")
  } else if(any(!num %in% all_nums)){
    stop("Invalid NUM input. The requested variable is not on the approved list.")
  }
  
  #cat:
  cat <-toupper(cat)
  
  if (length(cat)>1) {
    stop("Invalid CAT input. Only 1 variable permitted.")
  } else if(any(!cat %in% all_cats)){
    stop("Invalid CAT input. The requested variable is not on the approved list.")
  }
  
  #geo: 
  geo <- toupper(geo)
  
  if (length(geo) > 1){
    stop("Invalid GEO input. Only 1 geography variable may be provided.")
  } else if (!geo %in% c("REGION", "DIVISION", "STATE", "ST")){
    stop("Invalid GEO input. The requested variable is not on the approved geography list.")
  } else { 
    #Correct ST vs. STATE if entered incorrectly
    if(year %in% c(2021, 2022) & geo=="STATE") {
      geo <- "ST"
    } else if(year %in% c(2023, 2024) & geo=="ST") {
      geo <- "STATE"
    }
  }
  
  #load dictionary: used for subgeo checks and later cleanXXX functions
  dictionary <- readRDS(paste0("data/vars",as.character(year),".rds"))

  #subgeo:
  subgeo_phrase <- unlist(dictionary[[geo]]$values$item)
  subgeo_code   <- names(subgeo_phrase)
  
  if(is.null(subgeo)){
    #Set defaults when NULL input
    subgeo <- ifelse(geo=="REGION", "1", 
                ifelse(geo=="DIVISION", "2", 
                  ifelse(geo %in% c("STATE","ST"), "11", NULL) ))
  } else if(length(subgeo) > 1){
    stop("Invalid SUBGEO input. Specify 'all' or a single level of sub-geography.")
  } else if(toupper(as.character(subgeo)) %in% c("ALL","*")){
    #Force 'all' to coded value for URL
    subgeo <- "*"
  } else if(!as.character(subgeo) %in% c(subgeo_code, subgeo_phrase)){
    #Only accepts exact case/spelling of phrase or the 2-digit number code
  stop(paste0("Invalid SUBGEO input. Requires a valid code number/phrase for ",geo," geography."))
  } else if(as.character(subgeo) %in% subgeo_phrase){
    #Force phrase to coded value for URL -- NOT WORKING!!
    subgeo <- subgeo_code[which(subgeo_phrase==as.character(subgeo))] 
  } else {
    #Force to character
      subgeo <- as.character(subgeo)
  }
  
  ## RUN API PULL FUNCTION
  parsed <- getFromAPI(year, num, cat, geo, subgeo)

  ## RUN TIDY FUNCTION
  tdata  <- makeTidy(parsed)
  
  ## RUN CLEANING FUNCTIONS
  tdata <- cleanCat(tdata,dictionary,cat)
  tdata <- cleanNum(tdata,dictionary,num)
  tdata <- cleanGeo(tdata,dictionary,geo)
  
}

```

## Testing

Example data pull using the `getPUMSData()` function, default inputs:

```{r}
#test <- getPUMSData(year=2021) ##REVISIT lexical error in json text
```

Example data pull using the `getPUMSData()` function, custom inputs

```{r}
test <- getPUMSData(year=2024, num="JWAP", cat="HHL", geo="REGION")
str(test)

test <- getPUMSData(year=2024, num="JWAP", cat="HHL", geo="STATE", subgeo="10")
str(test)

test <- getPUMSData(year=2024, num="GASP", cat="HHL", geo="DIVISION", subgeo="all")
str(test)
```

Expected to fail (`getPUMSData`):

```{r}
#| error: true

#Invalid Nums
getPUMSData(num=c("AGEP", "GASP"))  #too many inputs
getPUMSData(num=c("PWGTP"))         #too few inputs
getPUMSData(num=c("STATE"))         #unapproved var

#Invalid Cats
getPUMSData(cat=c("SEX","FER"))  #too many inputs
getPUMSData(cat=c("STATE"))      #unapproved var

#Invalid Geo
getPUMSData(geo=c("STATE","REGION"))  #too many inputs
getPUMSData(geo=c("STA"))             #unapproved var

#Invalid Subgeo
getPUMSData(geo="STATE", subgeo="1")      #state codes are two digits
getPUMSData(geo="STATE", subgeo="Texas")  #state codes are two digits

#getPUMSData(geo="STATE", subgeo="Texas/TX")  #should work

```

Expected to fail (`getPUMSVars`):

```{r}
#| error: true

getPUMSVars(2020)  #Invalid year
getPUMSVars()      #No Year
```

# --- HERE ---

# Data Summarizing Functions

# Investigation

**Topic**

**Results**

**Discussion**
