---
title: "ST558 - Project 1"
author: "Kayla Nowak"
format: html
---

# Introduction

This project builds a set of generalized queries for accessing and summarizing data from the Public Use Microdata Sample (PUMS). PUMS is part of the US Census Bureau's annual American Community Survey.

-   Data Source: <https://www.census.gov/data/developers/data-sets/census-microdata-api.html>
-   Variables (from 2024): <https://api.census.gov/data/2024/acs/acs1/pums/variables.html>

This project was designed to access the 1-Year PUMS data sets for years 2021 through 2024 only. Additionally, PUMS data for Puerto Rico are not included the scope of this project.

# Data Processing Functions

Load essential packages and set API key.

```{r}
#| message: False
#| warning: False

library(tidyverse)
library(jsonlite)
library(ggplot2)
```

Core R objects, used throughout.

```{r}
my_key <- "07a32887514d391e8f71d7c20274286c27064b18"

all_years <- c("2021", "2022", "2023", "2024")

all_nums <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
all_cats <- c("FER", "HHL", "SCH", "SCHL", "SEX")
all_vars <- c("PWGTP", all_nums, all_cats, "ST", "STATE", "REGION", "DIVISION")
```

## Variable Values

Most PUMS data are reported as coded values, with codes defined in the online data dictionary for each year. Dictionaries (aka codebooks, or variables values documents) were downloaded to the repository to facilitate future access. The function used to perform this task is described below.

### `getPUMSVars()`

This function pulls and parses all applicable variable names and values for the PUMS data in a given year.

The function returns a single, year-specific .rda file, which is saved in the repository's `/data` folder for future access. Later functions reference these .rda files when defining the `dictionary`.

-   Input: `year` (2021, 2022, 2023, 2024)
-   Return: `data/vars_yyyy.rda` (as list)

```{r}

getPUMSVars <- function(year){
  
  yearc <- as.character(year)

  ## Define variables of interest
  if (yearc %in% c("2021", "2022")) {
    want <- all_vars[-14]
  } else if (yearc %in% c("2023", "2024")) {
    want <- all_vars[-13]
  } else {
    stop("Invalid year input. Function only accepts 2021-2024")
  }
  
  ## Pull + Parse JSON
  url <- paste0("https://api.census.gov/data/",yearc,"/acs/acs1/pums/variables.json")
  pulled  <- httr::GET(url)
  parsed  <- fromJSON(rawToChar(pulled$content))
  
  ## Subset List to Variables of Interest
  vars <- parsed$variables[want]
  
  ## Save to /data as .rds
  saveRDS(vars, file= paste0("data/vars",yearc,".rds"))
}

```

Called for all valid years, then commented out.

*Note: This code only needs to be run once since the results are permanently stored under in the repository under /data. Calls have been commented out for computational efficiency of future QMD executions.*

```{r}
#getPUMSVars(2021)
#getPUMSVars(2022)
#getPUMSVars(2023)
#getPUMSVars(2024)
```

## API Pull: Helpers

The following helper functions perform dedicated tasks and are called within the main, wrapper function (`getPUMSData` defined later).

*Note: testing for all helper functions are included in the "API Pull: Main" testing section.*

### `queryAPI()`

This function queries the Census API for PUMAS data and returns a matrix of the requested data.

-   Input: all query parameters, as given to the main/wrapper function.
-   Return: a matrix

```{r}
queryAPI <- function(year, nums, cats, geo, subgeo,...){
  
  q <- paste(c(cats, nums), collapse=",") 

  url <- paste0("https://api.census.gov/data/",as.character(year),
                "/acs/acs1/pums?get=",q,
                "&for=",tolower(geo),":",subgeo,
                "&key=", my_key)
  
  pulled <- httr::GET(url)
  parsed <- fromJSON(rawToChar(pulled$content))
}
```

### `makeTidy()`

This function converts the requested data to a tibble with meaningful column names.

-   Input: a matrix generated by `queryAPI`
-   Return: a tibble

```{r}
makeTidy <- function(parsed,...){

  #Get colnames from 1st row
  colnames(parsed) <- parsed[1,]
  
  #Make parsed tidy with colnames
  tdata  <- as_tibble(parsed) |> slice(-1)
}
```

### `cleanCat()`

This function converts the requested categorical variables to factors, based on the relevant data dictionary stored in `/data`.

-   Input: a tibble generated by `makeTidy`, the data dictionary, and the `cats` user input
-   Return: an updated tibble

```{r}
cleanCat <- function(tdata,dictionary,cats,...){

  for (i in 1:length(cats)){
    lookup <- unlist(dictionary[[cats[i]]]$values$item)
    tdata[,cats[i]] <- factor(tdata[[cats[i]]],     #tibble-to-vector syntax
                              levels=names(lookup), 
                              labels=lookup)
  }

 return(tdata)
}
```

### `cleanTime()`

This is a helper function, called within the `cleanNum` function defined below to convert any requested time variables (JWAP and JWDP) to interpretable numeric time data.

Raw time data are numeric codes representing a period of time (HH:MM am/pm to HH:MM am/pm). These values are decoded using the relevant data dictionary, and the midpoint of the time period is returned under the original variable name.

Midpoint time is calculated as the start time + half of the total time period. Numeric times are captured as HH:MM:SS on a 24-hour clock.

-   Input: a tibble generated by `makeTidy`, the data dictionary, and a time variable
-   Return: an updated tibble

```{r}
cleanTime <- function(tdata,dictionary,tvar){

  ## Load dictionary
  lookup <- unlist(dictionary[[tvar]]$values$item)
 
  ## Detailed cleaning (all local to function)
  local  <- tdata |>
    #Convert time codes to character strings, use NA for coded "N/A"
    mutate(tfmtz3 = stringr::str_pad(tdata[[tvar]], 3, pad="0"),
           tfmtz3 = ifelse(tfmtz3=="000", "0", tfmtz3),
           tdecod = factor(tfmtz3, levels=names(lookup), labels=lookup),
           tchar  = ifelse(str_detect(tdecod,"N/A"), NA, as.character(tdecod))) |>
    #Parse time windows into start (t1) and stop (t2) times, all character values
    separate_wider_delim(cols=tchar, delim=" to ", names=c("t1","t2")) |>
    #Convert character time to numeric time and calculate midpoint time
    mutate(t1 = lubridate::parse_date_time(gsub("\\.", "", t1), orders = "I:M p"),
           t2 = lubridate::parse_date_time(gsub("\\.", "", t2), orders = "I:M p")) |>
    #Calculate midpoint time, stored as time
    mutate(mpt = hms::as_hms(t1 + (t2-t1)/2))
 
  ## Replace original variable with new midpoint
  tdata[,tvar] <- local[,"mpt"]
  
  return(tdata)
}
```

### `cleanNum()`

This function converts the requested (true) numeric variables to interpretable numeric values, based on the relevant data dictionary stored in `/data`.

Coded "N/A" responses are converted to NA elements, and time periods (JWAP and JWDP) are converted to midpoint time. Midpoint time is stored as R numeric data with the 12-hour time format (`I:M p`).

-   Input: a tibble generated by `makeTidy`, the data dictionary, and the `nums` user input
-   Return: an updated tibble

```{r}
cleanNum <- function(tdata,dictionary,nums,...){

  #true numeric
  tdata <- tdata |>
      mutate(across(any_of(c("PWGTP","AGEP", "GASP", "GRPIP","JWMNP")), as.numeric))
  
  #fix coded NA
  if(any(nums=="GASP")){
    tdata <- tdata |> mutate(GASP = ifelse(GASP==3,NA,GASP)) #set coded missing to zero
  }
  
  #time cleaning
  if(any(nums=="JWAP")){
    tdata <- cleanTime(tdata,dictionary, tvar="JWAP")
  }
  
  if(any(nums=="JWDP")){
    tdata <- cleanTime(tdata,dictionary, tvar="JWDP")
  } 
  
  return(tdata)
} 
```

### `cleanGeo()`

This function converts the requested geographic variable to a factor based on the relevant data dictionary stored in `/data`.

-   Input: a tibble generated by `makeTidy`, the data dictionary, and the `geo` user input
-   Return: an updated tibble

```{r}
cleanGeo <- function(tdata,dictionary,geo,year,...){
  
  #Store variable labels
  geod <- ifelse(geo=="STATE" & year %in% c(2021,2022), "ST", geo)
  lookup <- unlist(dictionary[[geod]]$values$item)
  
  #Apply labels and levels
  if(geo %in% c("STATE", "ST")) {
    tdata <- tdata |> mutate(state = factor(state, levels=names(lookup), labels=lookup))
  } else if(geo=="REGION"){
    tdata <- tdata |> mutate(region = factor(region, levels=names(lookup), labels=lookup))
  } else if(geo=="DIVISION"){
    tdata <- tdata |> mutate(division = factor(division, levels=names(lookup), labels=lookup))
  }
  
  return(tdata)
}
```

### `getOneYear()`

This is a helper function that runs all prior helper functions. It is designed to query, tidy, and clean PUMS data for a single year. Within the main wrapper function (`getPUMSData` defined later), this helper function may be called multiple times to compile multiple years of data.

-   Input: all query parameters, as given to the main/wrapper function.
-   Return: a cleaned, single year tibble

```{r}
getOneYear <- function(year, nums, cats, geo, subgeo,...){

  ## CHECK SUBGEO AGAINST DICTIONARY
  dictionary <- readRDS(paste0("data/vars",as.character(year),".rds"))
  
  ##-----more to add
  
  ## RUN API PULL FUNCTION
  parsed <- queryAPI(year, nums, cats, geo, subgeo)

  ## RUN TIDY FUNCTION
  tdata  <- makeTidy(parsed)
  
  ## RUN CLEANING FUNCTIONS
  tdata <- cleanCat(tdata,dictionary,cats)
  tdata <- cleanNum(tdata,dictionary,nums)
  tdata <- cleanGeo(tdata,dictionary,geo,year)
  
    #Add year as a variable
  tdata <- tdata |> mutate(YEAR=factor(year, levels=c(2021, 2022, 2023, 2024)))
  
  #Assign class
  class(tdata) <- c("census", class(tdata))
  return(tdata)
}

```

## API Pull: Main

The following wrapper function performs all data querying, cleaning, and compiling operations for PUMS data. Multiple years may be requested, but all other arguments are restricted to a single value.

### `getPUMSData`

This function accepts multiple years and returns a single, combined tibble with cleaned data. Valid user inputs are as follows.

-   `years` accepts any combination of 2021, 2022, 2023, and/or 2024 (default = 2024)
-   `nums` accepts any of these numeric variables: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP (default = AGEP)
    -   Note: "PWGTP" is automatically included in addition to the requested variable
-   `cats` accepts any of these categorical variables: FER, HHL, SCH, SCHL, SEX (default = SEX).
-   `geo` accepts one level of geography: REGION, DIVISION, STATE (default = STATE)
    -   Note: either "STATE" or "ST" may be used to request state-level geography
-   `subgeo` accepts one sub-level of geography: ALL, NULL, or a numeric code
    -   ALL preserves all available levels of the requested geography (i.e., no subset applied).

    -   NULL subsets the data to a default sub-level of geography:

        -   1 ("Northeast") for REGION geography
        -   1 ("New England (Northeast region)") for DIVISION geography
        -   11 ("District of Columbia/DC") for STATE geography

    -   Numeric codes for all available sub-levels of geography can be found in the year-specific online data dictionary: [2021](https://api.census.gov/data/2021/acs/acs1/pums/variables.html), [2022](https://api.census.gov/data/2022/acs/acs1/pums/variables.html), [2023](https://api.census.gov/data/2023/acs/acs1/pums/variables.html), or [2024](https://api.census.gov/data/2024/acs/acs1/pums/variables.html).

```{r}
getPUMSData <- function(years=c(2024), nums=c("AGEP"), cats=c("SEX"), geo="STATE", subgeo=NULL,...){

  ## CHECK INPUTS
  
  #years:
  years <- as.numeric(years)
  
  if(any(!as.character(years) %in% c("2021", "2022", "2023", "2024"))){
    stop("Invalid YEAR input. Function only accepts 2021-2024.")    
  } else if (length(unique(years)) != length(years)){
    stop("Invalid YEAR input. Same year entered multiple times.")
  }
  
  #nums:
  nums <- unique(c("PWGTP",toupper(nums)))
  
  if (any(!nums %in% c("PWGTP",all_nums))){
    stop("Invalid NUMS input. At least 1 variable is not on the approved list.")
  } else if(length(nums)==1 && nums=="PWGTP"){
    stop("Invalid NUMS input. At lease 1 variable other than PWGTP is required.")
  }

  #cats:
  cats <- unique(toupper(cats))
  
  if (any(!cats %in% all_cats)){
    stop("Invalid CATS input. At least 1 variable is not on the approved list.")
  } 

  #geo:
  geo <- toupper(geo)

  if (length(geo) > 1){
    stop("Invalid GEO input. Only 1 geography variable may be provided.")
  } else if (!geo %in% c("REGION", "DIVISION", "STATE", "ST")){
    stop("Invalid GEO input. The requested variable is not on the approved geography list.")
  } else{
    geo <- ifelse(geo=="ST", "STATE", geo) 
  }
  
  #subgeo:
  #these are top-level checks, getOneYear checks year-specific dictionary compliance
  if(is.null(subgeo)){  
    #Set defaults
    subgeo <- ifelse(geo=="REGION", "1",
                ifelse(geo=="DIVISION", "1",
                  ifelse(geo %in% c("STATE","ST"), "11", NULL) ))
  } else if(length(subgeo) > 1){
    stop("Invalid SUBGEO input. Specify 'all' or a single level of sub-geography.")
  } else if(toupper(subgeo) %in% c("ALL","*")){
    subgeo <- "*"
  } else{
      subgeo <- as.character(subgeo)
  }
  
  #print final parameters to console
  message("Run for: ", paste(years,collapse="|"), " with ", paste(nums,collapse="|"), " & ", paste(cats,collapse="|"), " for ", geo, " at ", subgeo)
  
  
  ## PULL CLEAN DATA
  all_years <- getOneYear(years[1],nums,cats,geo,subgeo)
  
  if(length(years)>1){
     for (i in 2:length(years)){
        all_years <- all_years |> rbind(getOneYear(years[i],nums,cats,geo,subgeo))
     }
  }
  
  return(all_years)
}
```

# Data Summarizing Functions

This section defines census-class tibble updates for the generic `summary` and `plot` functions.

### `summary.census()`

The following code defines a generic `summary()` function for census class data. All summaries are weighted by PWGTP, and results are returned as a two-element list:

1.  `"mean_sd"` is a tibble containing weighted mean and standard deviation for each numeric variable (1 row per variable)
2.  `"count"` is a list of tibbles (1 tibble per variable) where each tibble contains the weighted count of each level of the respective categorical variable.

This function can be simply called by `summary(tdata)`, which automatically includes all available numeric and categorical variables in the tibble . Individual variable(s) can also be specified by user input (via `pick_nums=` and `pick_cats=`). All summaries are weighted by PWGTP.

Note: numeric time (JWAP and JWDP), the weighting factor (PWGTP), and levels of geography are *not* available to summarize with this function. If requested, these variables are automatically excluded (no warning message).

Note: any requested variables that are not present in the provided dataset are automatically excluded (no warning message).

-   Input: a census-class tibble generated by `getPUMSData`, numeric variables of interest (optional), categorical variables of interest (optional).
-   Return: a list with two elements ("mean_sd", "count")

```{r}
summary.census <- function(tdata, pick_nums="ALL", pick_cats="ALL"){

  ## PROCESS INPUTS
  pick_nums <- toupper(pick_nums)
  pick_cats <- toupper(pick_cats)

  valid_nums <- c("AGEP", "GASP", "GRPIP", "JWMNP")
  valid_cats <- c("YEAR", all_cats)

  #Refine pick_nums list
  if (pick_nums=="ALL"){
    pick_nums <- valid_nums
  } else if(all(!pick_nums %in% valid_nums)){
    stop("No valid inputs were provided for PICK_NUMS.")
  } 
  
  #Only continue with valid options that appear in the dataset
  pick_nums <- intersect(intersect(pick_nums, valid_nums), names(tdata))

  
  #Refine pick_cats list
  if (pick_cats=="ALL"){
    pick_cats <- valid_cats
  } else if(all(!pick_cats %in% valid_cats)){
    stop("No valid inputs were provided for PICK_CATS.")
  } 
  
  #Only continue with valid options that appear in the dataset
  pick_cats <- intersect(intersect(pick_cats, valid_cats), names(tdata))
 

  ## NUMERIC SUMMARY (MEAN, SD)
  just_means <- tdata |>
    select(PWGTP, any_of(pick_nums)) |>
    summarize(across(where(is.numeric),
                     .fns = list("mean" = ~ (sum(.x*PWGTP, na.rm=TRUE)/sum(PWGTP, na.rm=TRUE))),
                     .names = "{.col}")) |>
    select(-PWGTP) |>
    pivot_longer(cols=everything(), names_to = "Variable", values_to="mean")
  
  just_sd <- tdata |>
    select(PWGTP, any_of(pick_nums)) |>
    summarize(across(where(is.numeric),
                     .fns = list("sd" = ~ (sqrt(sum(.x**2*PWGTP, na.rm=TRUE)/sum(PWGTP, na.rm=TRUE)
                                                - (sum(.x*PWGTP, na.rm=TRUE)/sum(PWGTP, na.rm=TRUE))**2))),
                     .names = "{.col}")) |>
    select(-PWGTP) |>
    pivot_longer(cols=everything(), names_to = "Variable", values_to="sd") 
  
  
  element_meansd <- full_join(just_means, just_sd, by="Variable")
  
  
  ## CATEGORICAL SUMMARY (N x level)
  list_count <- vector("list", length(pick_cats))
  names(list_count) <- pick_cats

  for(i in 1:length(pick_cats)){
     element_count <- tdata |> count(tdata[,pick_cats[i]], wt=PWGTP)
     list_count[[i]] <- element_count
  }

 
  ## COMPILE RESULTS AS A LIST
  results <- list("mean_sd" = element_meansd,
                  "count" = list_count)

  return(results)
}

```

### `plot.census()`

The following code defines a generic `plot()` function for census class data.

This function creates a vertical box plot display for 1 numeric variable, grouped by 1 categorical variable. Exactly 1 variable of each type is required, and all summaries are weighted by PWGTP.

-   `num` accepts any 1 of these numeric variables: AGEP, GASP, GRPIP, JWMNP
-   `cat` accepts any 1 of these categorical variables: FER, HHL, SCH, SCHL, SEX, YEAR.

Note: numeric time (JWAP and JWDP), the weighting factor (PWGTP), and levels of geography are *not* available to summarize with this function. However, the level(s) of geography used to create the tibble are noted in the display.

-   Input: a census-class tibble generated by `getPUMSData`, 1 numeric variable of interest, categorical variables of interest.
-   Return: a vertical, grouped box plot

***axis label wrapping option?***

```{r}
plot.census <- function(tdata, num, cat){
  
  ## CHECK INPUTS
  if(is.null(num) | is.null(cat)){
    stop("Invalid input. Exactly 1 NUM and 1 CAT must be specified.")
  } else if(length(num) != 1 | length(cat) != 1){
    stop("Invalid input. Exactly 1 NUM and 1 CAT must be specified.")
  }
  
  num <- toupper(num)
  cat <- toupper(cat)
  if(!num %in% c("AGEP", "GASP", "GRPIP", "JWMNP")){
    stop("Invalid NUM input. The variable is not on the approved list.")
  } else if(!cat %in% c(all_cats, "YEAR")){
    stop("Invalid CAT input. The variable is not on the approved list.")
  }

  ## PLOT
  g <- ggplot(data=tdata,
              aes(x=get(cat), y=get(num), weight=PWGTP, fill=get(cat)) )

  g + geom_boxplot() +
    labs(x=cat, y=num) +
    scale_fill_discrete(name=cat) + 
    theme_minimal()
  
}
```

# Testing

## Correct Usage

Example function applications: default settings.

```{r}

#API Data Pull
default <- getPUMSData()   
str(default)

#Summary
summary(default)

#Graphics (AGEP x SEX)
plot(default, num="AGEP", cat="SEX")

```

Example function applications: multi-year data.

```{r}
#| eval: FALSE

#API Data Pull
test <- getPUMSData(years=c(2021,2022), nums= c("AGEP", "JWAP"), geo="DIVISION")
str(test)

#Summary
summary(test)

#Graphics (AGEP x YEAR)
plot(test, num="AGEP", cat="YEAR")

```

Example function applications: all covariates

```{r}
#| eval: FALSE

#API Data Pull
test <- getPUMSData(nums=all_nums, cats=call_cats, geo="REGION", subgeo="2")
str(test)

#Summary
summary(test)

#Graphics ( x )
plot(test, num="", cat="")
```

## Incorrect Usage

For `getPUMSVars`:

```{r}
#| error: true
getPUMSVars(2020)           #invalid year
```

For `getPUMSData`:

```{r}
#| eval: false
#| error: true

#Invalid Nums
getPUMSData(num=c("AGEP", "GASP"))  #too many inputs
getPUMSData(num=c("PWGTP"))         #too few inputs
getPUMSData(num=c("STATE"))         #unapproved variable

#Invalid Cats
getPUMSData(cat=c("SEX","FER"))     #too many inputs
getPUMSData(cat=c("STATE"))         #unapproved variable

#Invalid Geo
getPUMSData(geo=c("STATE","REGION"))  #too many inputs
getPUMSData(geo=c("STA"))             #unapproved variable

#Invalid Subgeo
#getPUMSData(geo="STATE", subgeo="1")      #state codes are two digits -- jk this works :)
#getPUMSData(geo="STATE", subgeo="Texas")  #crashes at parse -- TRY TO FIX
#getPUMSData(geo="DIVISION", subgeo="22")  #crasehs at parse -- TRY to FIX

#getPUMSData(geo="STATE", subgeo="Texas/TX")  #should work

```

For `summary.census`:

```{r}

```

For `plot.census`:

```{r}

```

# Investigation

**Topic**

**Results**

**Discussion**
