---
title: "ST558 - Project 1"
author: "Kayla Nowak"
format: html
---

## Introduction

This project builds a set of generalized queries for accessing and summarizing data from the Public Use Microdata Sample (PUMS). PUMS is part of the US Census Bureau's annual American Community Survey.

-   Data Source: <https://www.census.gov/data/developers/data-sets/census-microdata-api.html>
-   Variables (from 2024): <https://api.census.gov/data/2024/acs/acs1/pums/variables.html>

This project was designed to access the 1-Year PUMS data sets for years 2021 through 2024 only. Additionally, PUMS data for Puerto Rico are not included the scope of this project.

## Functions

Load essential packages and set API key.

```{r}
#| message: False
#| warning: False

library(tidyverse)
library(httr)
library(jsonlite)
library(ggplot2)

#allow errors to persist with message
```

Core R objects, used throughout.

```{r}
my_key <- "07a32887514d391e8f71d7c20274286c27064b18"

all_years <- c("2021", "2022", "2023", "2024")

all_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER",
              "HHL", "SCH", "SCHL", "SEX", "ST", "STATE", "REGION", "DIVISION")

```

### 1. Data Processing

**Variable Values**

Pull and parse variable names and values for the PUMS data in a given year.

-   Input: `year` (2021, 2022, 2023, 2024)
-   Return: `data/vars_yyyy.rda` (as list)

Function:

```{r}

getPUMSVars <- function(year){
  
  yearc <- as.character(year)

  ## Define variables of interest
  if (yearc %in% c("2021", "2022")) {
    want <- all_vars[-14]
  } else if (yearc %in% c("2023", "2024")) {
    want <- all_vars[-13]
  } else {
    stop("Invalid year input. Function only accepts 2021-2024")
  }
  
  ## Pull + Parse JSON
  url <- paste0("https://api.census.gov/data/",yearc,"/acs/acs1/pums/variables.json")
  pulled  <- GET(url)
  parsed  <- fromJSON(rawToChar(pulled$content))
  
  ## Subset List to Variables of Interest
  vars <- parsed$variables[want]
  
  ## Save to /data as .rds
  saveRDS(vars, file= paste0("data/vars",yearc,".rds"))
}

```

Called for all valid years:

```{r}
#getPUMSVars(2021)
#getPUMSVars(2022)
#getPUMSVars(2023)
#getPUMSVars(2024)
```

**API Data Pull**

The `getPUMSData` function ...

-   `year` accepts 2021, 2022, 2023, 2024
-   `nums` accepts up to two numeric variables: PWGTP, AGEP, GASP, GRPIP, JWAP, JWDP, and/or JWMNP
    -   if one variable is provided, "PWGTP" is the second variable
    -   if two variables are provided, "PWGTP" must be one of the two variables
    -   the default inputs are c("AGEP", "PWGTP")
-   `cats` accepts one categorical variable: FER, HHL, SCH, SCHL, SEX.
    -   the default input is "SEX"
- `geo` accepts one level of geography: REGION, DIVISION, STATE
    - the default input is "STATE"
    - note: either "STATE" or "ST" may be used to request state-level geography

```{r}
getPUMSData <- function(year, nums=c("AGEP","PWGTP"), cats=c("SEX"), geo=c("STATE"), subgeo, ...) {
  
  ## Check YEAR input
  yearc <- as.character(year)
  
  if (!yearc %in% c("2021", "2022", "2023", "2024")) {
    stop("Invalid year input. Function only accepts 2021-2024.")
  }
  
  ## Check NUMS input
  nums <- toupper(nums)
  if (length(nums) > 2 | (length(nums)==2 & !any(nums == "PWGTP")) ){
    stop("Invalid NUMS input. Only 1 numeric variable in addition to PWGTP may be requested.")
    
  } else if (any(!nums %in% c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"))) {
    stop("Invalid NUMS input. The requested variable(s) are not on the approved numeric list.")
  } else if (length(nums==1) & nums == "PWGTP"){
    stop("Invalid NUMS input. At least 1 numeric variable other than PWGTP must be provided.")
  } else if (length(nums==1) & nums != "PWGTP"){
    nums <- c(nums, "PWGTP")
  }
  
  
  ## Check CATS input
  cats <- toupper(cats)
  if (length(cats) > 1){
    stop("Invalid CATS input. Only 1 categorical variable may be provided.")
    
  } else if (!cats %in% c("FER", "HHL", "SCH", "SCHL", "SEX")){
    stop("Invalid CATS input. The requested variable is not on the approved categorical list.")
    
  }

  ## Check GEO input
  geo <- toupper(geo)
  if (length(geo) > 1){
    stop("Invalid GEO input. Only 1 geography variable may be provided.")
    
  } else if (!geo %in% c("REGION", "DIVISION", "STATE", "ST")){
    stop("Invalid GEO input. The requested variable is not on the approved geography list.")
    
  } else {
    
    #Force "ST" vs. "STATE" if entered wrong.
    if(yearc %in% c("2021", "2022") & geo=="STATE") {
      geo <- "ST"
    } else if(yearc %in% c("2023", "2024") & geo=="ST") {
      geo <- "STATE"
    }
  }
  
  ## Check REGION input
  
  ## Run API pull via helper function
  
}

```

### 2. Data Summarizing

## Testing

### getPUMSVars()

Expected to fail:

```{r}
#| error: true

getPUMSVars(2020)  #Invalid year
getPUMSVars()      #No Year
```

### getPUMSData()

Expected to work:


Expected to fail:

```{r}
#| error: true

#Invalid Year
getPUMSData(2020, nums, cats, geo, subgeo)

#Invalid Nums
getPUMSData(2021, nums=c("AGEP", "GASP"))  #too many inputs
getPUMSData(2021, nums=c("PWGTP"))         #too few inputs
getPUMSData(2021, nums=c("STATE"))         #unapproved var
getPUMSData(2021, nums=c("STATE","PWGTP")) #unapproved var

#Invalid Cats
getPUMSData(2021, nums=c("AGEP"), cats=c("SEX","FER"))  #too many inputs
getPUMSData(2021, nums=c("AGEP"), cats=c("STATE"))      #unapproved var

#Invalid Geo
getPUMSData(2021, nums=c("AGEP"), geo=c("STATE","REGION"))  #too many inputs
getPUMSData(2021, nums=c("AGEP"), geo=c("STA"))             #unapproved var

#Invalid Subgeo


```



## Investigation

**Topic**

**Results**

**Discussion**
